|

### 2021-04-07 02-Chp-Background.tex
End of day: 0

---
output:
  bookdown::pdf_document2:
    template: templates/brief_template.tex
  bookdown::word_document2: default
  bookdown::html_document2: default
documentclass: book
---

```{block type='savequote', include=knitr::is_latex_output(), quote_author='(ref:dara-quote)', echo = TRUE}
Science knows it doesn't know everything; otherwise, it'd stop.
```

(ref:dara-quote) --- Dara O'Briain

```{r, echo = FALSE, warning=FALSE, message=FALSE}
source("R/doc_options.R")
knitr::read_chunk("R/background.R")
```

# Background & Objectives {#background-heading}

\minitoc <!-- this will include a mini table of contents-->

## Lay summary

Around __[NUMBER]__ people in the UK live with dementia, and by 2040, nearly twice as many will have the condition. No cure for dementia exists, so researchers are trying to find ways to prevent the condition. This involves finding risk factors for dementia that we can change. Risk factors increase a person's risk of developing a disease. Avoiding a risk factor does not guarantee that a person will not develop dementia, but makes it less likely. A key risk factor for dementia may be the levels of lipids (fatty substances such as cholesterol) in a person's blood. However, researchers remain unsure, as studies disagree on whether a role for lipids in dementia exists.

@@

I will use all available evidence, including that preprinted by authors (i.e. not formally reviewed by others in their field to assess the accuracry of their claims) to assess whether lipid levels cause dementia.

@@

<!----------------------------------------------------------------------------->

## Introduction

<!-- This chapter provides the context of the thesis. It briefly gives the background on the problem of increased adiposity, the biology of adipose tissue, and diseases linking increased body mass index and mortality. The chapter briefly explores different measures of increased adiposity and potential hypotheses for underlying aetiology of diseases associated with increased adiposity. Finally, the potential application of metabolites and Mendelian randomization in the context of understanding underlying aetiology are discussed. *From Matt Lee's thesis.* -->

@@

<!----------------------------------------------------------------------------->

## Dementia

@@

The most common types of which are Alzheimer's disease (AD) and vascular dementia (VaD) [1].

@@

Dementia is major neurocognitive disorder, with symptoms including impairment of executive cognitive functions such as speech, judgement and memory. The most common causes of dementia are Alzheimer's disease and vascular dementia, accounting for ~60-80% and ~10% of cases respectively. The remaining 10-30% of cases are caused other dementia subtypes (e.g. Lewy Body dementia) or by progression of other neurological diseases (e.g. Parkinson's disease).

@@

Dementia is difficult to diagnose, primarily due to the absence of a gold standard test for the condition. <!-- is this true, or would it be better to say that there is no gold standard test without an autopsy?--> Information from multiple diagnostic tools are utilised, from medical history examination, through assessment of patients mental ability (e.g. the Mini-Mental State Examination), to clinical tests (e.g. Magnetic Resonance Imaging (MRI) scans).

### Public Health importance

Dementia is quickly becoming a critically important public health issue. Despite the age-specific incidence and prevalence of dementia remaining relatively constant over time,[@prince2016] an ageing population looks set to create a dementia epidemic, particularly in Westernised countries. While approximately have received a dementia diagnosis, the true number of people currently living with dementia in the UK is thought to be closer to 850,000, with this figure expected to double by 2040.[@baker2019] Globally, the prevalence of dementia is expected to reach 75 million by 2030.[@prince2016] Despite the increasing number of dementia cases and decades of research, there remains much unknown about the pathogenesis and progression of the disease.

### Chronic underfunding

@@

<!----------------------------------------------------------------------------->

### Natural history

Much remains unknown about dementia pathogenesis, despite research implicating amyloid plaques and neurofibrillary tangles as potential mechanisms of disease [2]. Progression from onset is an average

<!----------------------------------------------------------------------------->

### Alzheimer's disease

@@

### Vascular dementia

@@

Diagnosed using the NINCDS-AIREN criteria.[@roman1993vascular]

<!----------------------------------------------------------------------------->

### Diagnostis criteria

@@

<!----------------------------------------------------------------------------->

```{r diagnosticCriteria-setup, results="hide", message=FALSE, echo = FALSE}
```

```{r diagnosticCriteria-table, message=FALSE, results="asis", echo = FALSE}
```

<!----------------------------------------------------------------------------->

Several sets of criteria are commonly used to diagnose dementia, and the underlying diseases for research purposes. Two of the most commonly used include the criteria of the Diagnostic and Statistical Manual of Mental Disorders, fourth edition (DSM-IV-TR)1 and the National Institute of Neurological Disorders and Stroke--Alzheimer Disease and Related Disorders (NINCDS-ADRDA).[@dubois2007]

[**Need first NINCDS-ADRDA citation and DSM-IV citations**]

@@

<!-- Note important differences between the scales and whether they affect results. Could cross-reference with an exploratory analysis in systematic review -->

@@

### Differing risk by sex

@@

Rate of dementia/AD differs in men and women <!--- Though is this just due to the different lifestyles and life expectancies --->

Previous studies have hinted that the relationship is dependent on individual-level characteristics such as age and sex [16,28].

<!----------------------------------------------------------------------------->

## Challenges in the study of dementia using traditional study designs

@@

Historically, studies on dementia have faced a range of challenges. As a range of determinants (genetic, environmental and lifestyle) are thought to jointly influence the risk, progression and outcomes of dementia, any individual association is likely to have only a small effect. Therefore, the statistical power need to detect these associations will often require sample sizes that are unfeasible using primary data collection. This fact is further complicated by the long latency of dementia, which necessitates a costly long-term approach to patient follow-up. Furthermore, dementia studies often have limited generalisability to the target population, as certain subgroups (e.g. the very old) are frequently under-represented in study cohorts due to difficulties associated with their recruitment.9 Finally, studies employing a case-control design are further limited by the high potential for differential recall bias between those who have and have not developed dementia.10

@@

*Cognitive impairment not dementia* (CIND) is another term used to described those with cognitive impairment but who fall below the diagnostic criteria for dementia.

@@

<!-- I -->

<!-- Consider sex as a biological variable -->

<!----------------------------------------------------------------------------->

### Economic impact

@@

Patients with dementia stay in hospital for longer,[@mollers2019]

@@

<!----------------------------------------------------------------------------->

### Treatments

@@

<!----------------------------------------------------------------------------->

### Risk factors and Preventative measures

@@

<!----------------------------------------------------------------------------->

## Serum lipids

@@

<!----------------------------------------------------------------------------->

### Range of lipids

@@

Range of other higher and lower density (VLDL, etc)

<!--- Will need to explain why I am focusing on these fractions rather that looking at others include VLDL, etc --->

<!----------------------------------------------------------------------------->

### Total cholesterol

@@

<!--- Include formula for working out total cholesterol and a table detailing the acceptable ranges. --->

@@

<!----------------------------------------------------------------------------->

### Triglycerides

@@

<!----------------------------------------------------------------------------->

### HDL-c

@@

<!----------------------------------------------------------------------------->

### LDL-c

LDL-c levels are not directly measured but instead are estimated from using the Friedewald formula,[@friedewald1972] where the:

$$LDLc \approx TC - HDLc - kTG$$ where $k$ is 0.20 if measurements are in mg/dl and 0.45 if in mmol/l.

@@

<!-- Measurement error is a big deal for this thesis, as will discuss measurement error regarding how  different codelists/measurement of AD vs vascular dementia affect results in later chapters. -->

### Statins

@@

<!----------------------------------------------------------------------------->

(ref:statin-mechanisam-cap) **Statin mechanism of action:**

(ref:statin-mechanisam-scap) Statin mechanism of action

```{r statin-mechanisam, echo = FALSE, results="asis", fig.cap='(ref:statin-mechanisam-cap)', out.width='100%', fig.scap='(ref:statin-mechanisam-scap)'}
knitr::include_graphics(file.path("figures/sys-rev/prisma_flow.png"))
```

<!----------------------------------------------------------------------------->

#### Lipophilic

@@

#### Hydrophilic

### Other lipid regulating agents (LRA)

@@

<!----------------------------------------------------------------------------->

## Evidence synthesis

From a brief scoping literature search, Systematic reviews of the literature exist, with two main caveats: __[CITATION NEEDED - REFER to MSc project]__

@@

<!----------------------------------------------------------------------------->

## Thesis focus: Integrating diverse sources of evidence

@@

While this thesis does include a research tool Chapter \@ref(sys-rev-tools) and a primary evidence generation element (Chapter 5 - observational analysis of CPRD data), this was performed with the intention of providing a further source of evidence for the evidence synthesis/triangulation aspect of the thesis.

More specifically this includes the accurate searching of sources of grey literature (as enabled by the too presented in Chapter \@ref(sys-rev-tools-heading)) and risk of bias assessment.

<!----------------------------------------------------------------------------->

### Study designs and sources of bias

<!-- Need to talk about biases here? Or have a different section: Diverse sources of evidence: different underlying biases -->

@@

Indeed in the "evidence pyramid", meta-analysis of trials sit proudly on top. However, there included several potential limitations to this trial-only approach.

RCTs may be unfeasible if the outcome of interest is one with a long prodomal period, such as dementia, <!--- Cross link here to section on presentation of dementia ---> as they would require exteremely long and costly follow-up.[@ritchie2015] <!--- Check citation --->

@@

It is no surprise then that the two know existing trials of statins for the prevention of dementia are in fact trials of statins for the prevention of coronary related outcomes.[@mcguinness2016a] The Prospective Study of Pravastatin in the Elderly (PROSPER) examined the protective effect of pravastating on coronary outcomes,

@@

Firstly, there was no clinical cognitive evaluation of patients to . In fact, the PROSPER trial reported not on dementia outcomes, but on the change in cognitive scores over a mean of 3.2 years. As highlighted in Section \@() <!--- Cross with section on how scores alone are insufficient --->, change in scores alone is insufficient to diagnose a dementia outcome.

@@

<!-- > Two major clinical trials are often cited as providing evidence that statins do not have an effect on the incidence ofdementia: the Prospective Study of Pravastatin in the Elderly (PROSPER) and the Medical Research Council/British Heart Foundation Heart Protection Study; however, because of methodologic limitations in relation to dementia outcomes in these two trials, the results of these trials are difficult to evaluate. Dementia incidence or cognitive outcomes were not preplanned endpoints in either of them, neither study included a clinical cognitive evaluation, and numbers of patients with follow-up information for cognitive evaluations were not reported in either study manuscript. In PROSPER a post hoc analysis compared changes in cognitive scores over a 3-year period between statintreated and placebo patients and found no significant differences. In the MRC/BHF HPS trial,27 similar percentages of participants (0.3% in each---statin vs placebo---group) developed dementia during the 5-year follow-up period. The report did not state how the outcome of dementia was determined (e.g., reported as an adverse event or by follow-up phone interview). *Taken from Cramer 2008 - included in review* -->

__Maybe take a stand here - and in the systematic review, and say that you are not going to __

There exist only two large scale randomized controlled trials of the effect of statins on dementia risk [CITE] - in fact these studies are the only two trials included in a 2016 review of statins for the prevention of dementia produced by the Cochrane [@mcguinness2016b]. Both showed no effect of

@@

A role for lipids in the aetiology of dementia is supported by both genetic linkage studies and functional cell biology studies. The generation of the amyloid plaques found in the brains of Alzheimer's patients is cholesterol dependent [CITE], while the most established genetic risk factor for late-onset dementia, apolipoprotein E (ApoE), is involved in cerebral cholesterol transport. Several other genes involved in cholesterol transport have also been found to be robustly associated with increased AD susceptibility [CITE].

@@

Several observational studies have examined the relationships between concentrations of serum lipids (total cholesterol (TC), low density lipoprotein cholesterol (LDL-c), high density lipoprotein cholesterol (HDL-c) and triglycerides) and both Alzheimer's disease and vascular dementia and reported widely different results. A high serum cholesterol concentration has been found to be associated with an increase in susceptibility to AD [8-13], though other studies have shown no association [14-17], or a reduced susceptibility [18,19]. With regards VaD, decreased levels of HDL-c appear to be associated with increased risk [18,20,21], while for LDL-c, studies have reported both positive and negative association [18,22].

Observational studies do not represent the only source of evidence examining the link between lipid levels and dementia. A recent Cochrane review identified two large randomised controlled trials (RCTs) of statins for the prevention of dementia, indicating that assignment to late-life lipid modifying treatment, as a proxy for lower lipid levels, does not reduce dementia risk [23]. However, both RCTs were limited, primarily by the relatively short follow-up period examined, while one failed to report the criteria used to diagnose dementia [24,25]. Finally, as they included only patients at high vascular risk, their generalisability to other settings is limited [23].

@@

The use of newer methodological approaches, such as Mendelian randomisation (MR), has also produced contradictory results. In brief, MR uses genetic variants that are both strongly associated with the exposure of interest and are independent from potential confounders to strengthen causal inference [26]. A recent MR study indicated that low levels of LDL-c may cause a reduction in AD risk [27]. However, this study was criticised as it did not exclude the region surrounding the ApoE gene, meaning the risk reduction observed could be driven by variants in this region.

n short, each individual epidemiological approach taken to examine this research question is subject to distinct biases and short-comings that limit our ability to infer a causal relationship between blood lipids and subsequent dementia risk. However, as each approach will be subject to distinct biases, these differences in design can be advantageous. If all approaches point, or triangulate, towards the same answer, this strengthens the evidence of a causal link between lipid levels and dementia risk [26]. In this context, systematically identifying, assessing and integrating all available evidence in a triangulation framework, regardless of study design or approach taken, may help to increase our confidence that a causal relationship truly exists.

@@

<!----------------------------------------------------------------------------->

### Different sources of bias

@@

The term "observational study", unlike RCT, belies wide range of analytical approaches,[@thiese2014] each with it's own limitations which can

@@

#### Misclassification bias

@@

<!--- Be careful here, as you don't account for misclassification in your CPRD study --->

@@

#### Reverse causation

Reverse causation is where the assumed exposure in an analytical framework is actually the outcome. **[CITATION NEEDED]** In the case of lipids and dementia, a reverse casaul relationship would be observed by the early prodomal effects of dementia influencing a change

#### Ascertainment bias

@@

<!--- This is a key advantage of this study, as we account for . --->

@@

See Appendix \@ref

&nbsp;

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% BRIEF VERSION OF OXFORD THESIS TEMPLATE FOR CHAPTER PREVIEWS

%%%%% CHOOSE PAGE LAYOUT
% format for PDF output (ie equal margins, no extra blank pages):
\documentclass[a4paper,nobind]{templates/ociamthesis}

% UL 30 Nov 2018 pandoc puts lists in 'tightlist' command when no space between bullet points in Rmd file
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{float}

% UL 1 Dec 2018, fix to include code in shaded environments
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

%UL 2 Dec 2018 add a bit of white space before and after code blocks
\renewenvironment{Shaded}
{
  \vspace{4pt}%
  \begin{snugshade}%
}{%
  \end{snugshade}%
  \vspace{4pt}%
}
%UL 2 Dec 2018 reduce whitespace around verbatim environments
\usepackage{etoolbox}
\makeatletter
\preto{\@verbatim}{\topsep=0pt \partopsep=0pt }
\makeatother

%UL 28 Mar 2019, enable strikethrough
\usepackage[normalem]{ulem}

%UL 3 Nov 2019, avoid mysterious error from not having hyperref included
\usepackage{hyperref}

%%%%% SELECT YOUR DRAFT OPTIONS
% Three options going on here; use in any combination.  But remember to turn the first two off before
% generating a PDF to send to the printer!

% This adds a "DRAFT" footer to every normal page.  (The first page of each chapter is not a "normal" page.)

% This highlights (in blue) corrections marked with (for words) \mccorrect{blah} or (for whole
% paragraphs) \begin{mccorrection} . . . \end{mccorrection}.  This can be useful for sending a PDF of
% your corrected thesis to your examiners for review.  Turn it off, and the blue disappears.

%%%%% BIBLIOGRAPHY SETUP
% Note that your bibliography will require some tweaking depending on your department, preferred format, etc.
% The options included below are just very basic "sciencey" and "humanitiesey" options to get started.
% If you've not used LaTeX before, I recommend reading a little about biblatex/biber and getting started with it.
% If you're already a LaTeX pro and are used to natbib or something, modify as necessary.
% Either way, you'll have to choose and configure an appropriate bibliography format...

% The science-type option: numerical in-text citation with references in order of appearance.
% \usepackage[style=numeric-comp, sorting=none, backend=biber, doi=false, isbn=false]{biblatex}
% \newcommand*{\bibtitle}{References}

% The humanities-type option: author-year in-text citation with an alphabetical works cited.
% \usepackage[style=authoryear, sorting=nyt, backend=biber, maxcitenames=2, useprefix, doi=false, isbn=false]{biblatex}
% \newcommand*{\bibtitle}{Works Cited}

%UL 3 Dec 2018: set this from YAML in index.Rmd
\usepackage[style=numeric-comp, sorting=none, backend=biber, doi=false, isbn=false]{biblatex}
\newcommand*{\bibtitle}{References}

% This makes the bibliography left-aligned (not 'justified') and slightly smaller font.
\renewcommand*{\bibfont}{\raggedright\small}

% Change this to the name of your .bib file (usually exported from a citation manager like Zotero or EndNote).
\addbibresource{}

\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{CSLReferences}%
  {}%
  {\par}

\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}

\usepackage{calc}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

%%%%% YOUR OWN PERSONAL MACROS
% This is a good place to dump your own LaTeX macros as they come up.

% To make text superscripts shortcuts
	\renewcommand{\th}{\textsuperscript{th}} % ex: I won 4\th place
	\newcommand{\nd}{\textsuperscript{nd}}
	\renewcommand{\st}{\textsuperscript{st}}
	\newcommand{\rd}{\textsuperscript{rd}}

%%%%% THE ACTUAL DOCUMENT STARTS HERE
\begin{document}

%%%%% CHOOSE YOUR LINE SPACING HERE
% This is the official option.  Use it for your submission copy and library copy:
\setlength{\textbaselineskip}{22pt plus2pt}
% This is closer spacing (about 1.5-spaced) that you might prefer for your personal copies:
%\setlength{\textbaselineskip}{18pt plus2pt minus1pt}

% UL: You can set the general paragraph spacing here - I've set it to 2pt (was 0) so
% it's less claustrophobic
\setlength{\parskip}{2pt plus 1pt}

% Leave this line alone; it gets things started for the real document.
\setlength{\baselineskip}{\textbaselineskip}

% all your chapters and appendices will appear here
\begin{savequote}
Science knows it doesn't know everything; otherwise, it'd stop.
\end{savequote}

\hypertarget{background-heading}{%
\chapter{Background \& Objectives}\label{background-heading}}

\minitoc

\hypertarget{lay-summary}{%
\section{Lay summary}\label{lay-summary}}

Around \textbf{{[}NUMBER{]}} people in the UK live with dementia, and by 2040, nearly twice as many will have the condition. No cure for dementia exists, so researchers are trying to find ways to prevent the condition. This involves finding risk factors for dementia that we can change. Risk factors increase a person's risk of developing a disease. Avoiding a risk factor does not guarantee that a person will not develop dementia, but makes it less likely. A key risk factor for dementia may be the levels of lipids (fatty substances such as cholesterol) in a person's blood. However, researchers remain unsure, as studies disagree on whether a role for lipids in dementia exists.

@@

I will use all available evidence, including that preprinted by authors (i.e.~not formally reviewed by others in their field to assess the accuracry of their claims) to assess whether lipid levels cause dementia.

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

\hypertarget{dementia}{%
\section{Dementia}\label{dementia}}

@@

The most common types of which are Alzheimer's disease (AD) and vascular dementia (VaD) {[}1{]}.

@@

Dementia is major neurocognitive disorder, with symptoms including impairment of executive cognitive functions such as speech, judgement and memory. The most common causes of dementia are Alzheimer's disease and vascular dementia, accounting for \textasciitilde60-80\% and \textasciitilde10\% of cases respectively. The remaining 10-30\% of cases are caused other dementia subtypes (e.g.~Lewy Body dementia) or by progression of other neurological diseases (e.g.~Parkinson's disease).

@@

Dementia is difficult to diagnose, primarily due to the absence of a gold standard test for the condition. Information from multiple diagnostic tools are utilised, from medical history examination, through assessment of patients mental ability (e.g.~the Mini-Mental State Examination), to clinical tests (e.g.~Magnetic Resonance Imaging (MRI) scans).

\hypertarget{public-health-importance}{%
\subsection{Public Health importance}\label{public-health-importance}}

Dementia is quickly becoming a critically important public health issue. Despite the age-specific incidence and prevalence of dementia remaining relatively constant over time,{[}@prince2016{]} an ageing population looks set to create a dementia epidemic, particularly in Westernised countries. While approximately have received a dementia diagnosis, the true number of people currently living with dementia in the UK is thought to be closer to 850,000, with this figure expected to double by 2040.{[}@baker2019{]} Globally, the prevalence of dementia is expected to reach 75 million by 2030.{[}@prince2016{]} Despite the increasing number of dementia cases and decades of research, there remains much unknown about the pathogenesis and progression of the disease.

\hypertarget{chronic-underfunding}{%
\subsection{Chronic underfunding}\label{chronic-underfunding}}

@@

\hypertarget{natural-history}{%
\subsection{Natural history}\label{natural-history}}

Much remains unknown about dementia pathogenesis, despite research implicating amyloid plaques and neurofibrillary tangles as potential mechanisms of disease {[}2{]}. Progression from onset is an average

\hypertarget{alzheimers-disease}{%
\subsection{Alzheimer's disease}\label{alzheimers-disease}}

@@

\hypertarget{vascular-dementia}{%
\subsection{Vascular dementia}\label{vascular-dementia}}

@@

Diagnosed using the NINCDS-AIREN criteria.{[}@roman1993vascular{]}

\hypertarget{diagnostis-criteria}{%
\subsection{Diagnostis criteria}\label{diagnostis-criteria}}

@@

\begin{table}[!h]

\caption{\label{tab:diagnosticCriteria-table}Diagnostic criteria used to diagnose different types of dementia}
\centering
\begin{tabular}[t]{lll}
\toprule
Disease & Criteria & Summary\\
\midrule
AD & NINCDS-ADRDA & Lorem ipsum\\
VaD & NINCDS-AIREN & Lorem ipsum 2\\
\bottomrule
\end{tabular}
\end{table}

Several sets of criteria are commonly used to diagnose dementia, and the underlying diseases for research purposes. Two of the most commonly used include the criteria of the Diagnostic and Statistical Manual of Mental Disorders, fourth edition (DSM-IV-TR)1 and the National Institute of Neurological Disorders and Stroke--Alzheimer Disease and Related Disorders (NINCDS-ADRDA).{[}@dubois2007{]}

{[}\textbf{Need first NINCDS-ADRDA citation and DSM-IV citations}{]}

@@

@@

\hypertarget{differing-risk-by-sex}{%
\subsection{Differing risk by sex}\label{differing-risk-by-sex}}

@@

Rate of dementia/AD differs in men and women

Previous studies have hinted that the relationship is dependent on individual-level characteristics such as age and sex {[}16,28{]}.

\hypertarget{challenges-in-the-study-of-dementia-using-traditional-study-designs}{%
\section{Challenges in the study of dementia using traditional study designs}\label{challenges-in-the-study-of-dementia-using-traditional-study-designs}}

@@

Historically, studies on dementia have faced a range of challenges. As a range of determinants (genetic, environmental and lifestyle) are thought to jointly influence the risk, progression and outcomes of dementia, any individual association is likely to have only a small effect. Therefore, the statistical power need to detect these associations will often require sample sizes that are unfeasible using primary data collection. This fact is further complicated by the long latency of dementia, which necessitates a costly long-term approach to patient follow-up. Furthermore, dementia studies often have limited generalisability to the target population, as certain subgroups (e.g.~the very old) are frequently under-represented in study cohorts due to difficulties associated with their recruitment.9 Finally, studies employing a case-control design are further limited by the high potential for differential recall bias between those who have and have not developed dementia.10

@@

\emph{Cognitive impairment not dementia} (CIND) is another term used to described those with cognitive impairment but who fall below the diagnostic criteria for dementia.

@@

\hypertarget{economic-impact}{%
\subsection{Economic impact}\label{economic-impact}}

@@

Patients with dementia stay in hospital for longer,{[}@mollers2019{]}

@@

\hypertarget{treatments}{%
\subsection{Treatments}\label{treatments}}

@@

\hypertarget{risk-factors-and-preventative-measures}{%
\subsection{Risk factors and Preventative measures}\label{risk-factors-and-preventative-measures}}

@@

\hypertarget{serum-lipids}{%
\section{Serum lipids}\label{serum-lipids}}

@@

\hypertarget{range-of-lipids}{%
\subsection{Range of lipids}\label{range-of-lipids}}

@@

Range of other higher and lower density

\hypertarget{total-cholesterol}{%
\subsection{Total cholesterol}\label{total-cholesterol}}

@@

@@

\hypertarget{triglycerides}{%
\subsection{Triglycerides}\label{triglycerides}}

@@

\hypertarget{hdl-c}{%
\subsection{HDL-c}\label{hdl-c}}

@@

\hypertarget{ldl-c}{%
\subsection{LDL-c}\label{ldl-c}}

LDL-c levels are not directly measured but instead are estimated from using the Friedewald formula,{[}@friedewald1972{]} where the:

\[LDLc \approx TC - HDLc - kTG\] where \(k\) is 0.20 if measurements are in mg/dl and 0.45 if in mmol/l.

@@

\hypertarget{statins}{%
\subsection{Statins}\label{statins}}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{figures/sys-rev/prisma_flow} \caption[Statin mechanism of action]{\textbf{Statin mechanism of action:}}\label{fig:statin-mechanisam}
\end{figure}

\hypertarget{lipophilic}{%
\subsubsection{Lipophilic}\label{lipophilic}}

@@

\hypertarget{hydrophilic}{%
\subsubsection{Hydrophilic}\label{hydrophilic}}

\hypertarget{other-lipid-regulating-agents-lra}{%
\subsection{Other lipid regulating agents (LRA)}\label{other-lipid-regulating-agents-lra}}

@@

\hypertarget{evidence-synthesis}{%
\section{Evidence synthesis}\label{evidence-synthesis}}

From a brief scoping literature search, Systematic reviews of the literature exist, with two main caveats: \textbf{{[}CITATION NEEDED - REFER to MSc project{]}}

@@

\hypertarget{thesis-focus-integrating-diverse-sources-of-evidence}{%
\section{Thesis focus: Integrating diverse sources of evidence}\label{thesis-focus-integrating-diverse-sources-of-evidence}}

@@

While this thesis does include a research tool Chapter \ref{sys-rev-tools} and a primary evidence generation element (Chapter 5 - observational analysis of CPRD data), this was performed with the intention of providing a further source of evidence for the evidence synthesis/triangulation aspect of the thesis.

More specifically this includes the accurate searching of sources of grey literature (as enabled by the too presented in Chapter \ref{sys-rev-tools-heading}) and risk of bias assessment.

\hypertarget{study-designs-and-sources-of-bias}{%
\subsection{Study designs and sources of bias}\label{study-designs-and-sources-of-bias}}

@@

Indeed in the ``evidence pyramid'', meta-analysis of trials sit proudly on top. However, there included several potential limitations to this trial-only approach.

RCTs may be unfeasible if the outcome of interest is one with a long prodomal period, such as dementia, as they would require exteremely long and costly follow-up.{[}@ritchie2015{]}

@@

It is no surprise then that the two know existing trials of statins for the prevention of dementia are in fact trials of statins for the prevention of coronary related outcomes.{[}@mcguinness2016a{]} The Prospective Study of Pravastatin in the Elderly (PROSPER) examined the protective effect of pravastating on coronary outcomes,

@@

Firstly, there was no clinical cognitive evaluation of patients to . In fact, the PROSPER trial reported not on dementia outcomes, but on the change in cognitive scores over a mean of 3.2 years. As highlighted in Section @() , change in scores alone is insufficient to diagnose a dementia outcome.

@@

\textbf{Maybe take a stand here - and in the systematic review, and say that you are not going to }

There exist only two large scale randomized controlled trials of the effect of statins on dementia risk {[}CITE{]} - in fact these studies are the only two trials included in a 2016 review of statins for the prevention of dementia produced by the Cochrane {[}@mcguinness2016b{]}. Both showed no effect of

@@

A role for lipids in the aetiology of dementia is supported by both genetic linkage studies and functional cell biology studies. The generation of the amyloid plaques found in the brains of Alzheimer's patients is cholesterol dependent {[}CITE{]}, while the most established genetic risk factor for late-onset dementia, apolipoprotein E (ApoE), is involved in cerebral cholesterol transport. Several other genes involved in cholesterol transport have also been found to be robustly associated with increased AD susceptibility {[}CITE{]}.

@@

Several observational studies have examined the relationships between concentrations of serum lipids (total cholesterol (TC), low density lipoprotein cholesterol (LDL-c), high density lipoprotein cholesterol (HDL-c) and triglycerides) and both Alzheimer's disease and vascular dementia and reported widely different results. A high serum cholesterol concentration has been found to be associated with an increase in susceptibility to AD {[}8-13{]}, though other studies have shown no association {[}14-17{]}, or a reduced susceptibility {[}18,19{]}. With regards VaD, decreased levels of HDL-c appear to be associated with increased risk {[}18,20,21{]}, while for LDL-c, studies have reported both positive and negative association {[}18,22{]}.

Observational studies do not represent the only source of evidence examining the link between lipid levels and dementia. A recent Cochrane review identified two large randomised controlled trials (RCTs) of statins for the prevention of dementia, indicating that assignment to late-life lipid modifying treatment, as a proxy for lower lipid levels, does not reduce dementia risk {[}23{]}. However, both RCTs were limited, primarily by the relatively short follow-up period examined, while one failed to report the criteria used to diagnose dementia {[}24,25{]}. Finally, as they included only patients at high vascular risk, their generalisability to other settings is limited {[}23{]}.

@@

The use of newer methodological approaches, such as Mendelian randomisation (MR), has also produced contradictory results. In brief, MR uses genetic variants that are both strongly associated with the exposure of interest and are independent from potential confounders to strengthen causal inference {[}26{]}. A recent MR study indicated that low levels of LDL-c may cause a reduction in AD risk {[}27{]}. However, this study was criticised as it did not exclude the region surrounding the ApoE gene, meaning the risk reduction observed could be driven by variants in this region.

n short, each individual epidemiological approach taken to examine this research question is subject to distinct biases and short-comings that limit our ability to infer a causal relationship between blood lipids and subsequent dementia risk. However, as each approach will be subject to distinct biases, these differences in design can be advantageous. If all approaches point, or triangulate, towards the same answer, this strengthens the evidence of a causal link between lipid levels and dementia risk {[}26{]}. In this context, systematically identifying, assessing and integrating all available evidence in a triangulation framework, regardless of study design or approach taken, may help to increase our confidence that a causal relationship truly exists.

@@

\hypertarget{different-sources-of-bias}{%
\subsection{Different sources of bias}\label{different-sources-of-bias}}

@@

The term ``observational study'', unlike RCT, belies wide range of analytical approaches,{[}@thiese2014{]} each with it's own limitations which can

@@

\hypertarget{misclassification-bias}{%
\subsubsection{Misclassification bias}\label{misclassification-bias}}

@@

@@

\hypertarget{reverse-causation}{%
\subsubsection{Reverse causation}\label{reverse-causation}}

Reverse causation is where the assumed exposure in an analytical framework is actually the outcome. \textbf{{[}CITATION NEEDED{]}} In the case of lipids and dementia, a reverse casaul relationship would be observed by the early prodomal effects of dementia influencing a change

\hypertarget{ascertainment-bias}{%
\subsubsection{Ascertainment bias}\label{ascertainment-bias}}

@@

@@

See Appendix @ref

~
